---
title: Data Preparation
author: "Sahil Thapar"
date: "12/9/2016"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

```{r, echo = F}
library("knitr") # To enable usage of R code chunks from ecternal file in the Rmd
library("RSQLite") # Connecting to RSQLite
library("dplyr")   # Working 
library("readr")   # Reading and writing files
library("tidyverse") # For all the tidying functions from the tidy universe
library("tibble")    # For working with tibbles
library('stringr')   # Working with strings and regex
library('lubridate') # Working with dates
library('forcats') # Working with factors
library('XML') # Parsing XML data
```

This section describes the data in depth and pre-processing.

### About the data

The data used is provided through
[Kaggle datasets](https://www.kaggle.com/hugomathien/soccer).

The data gives detailed information on matches played across 11 different
leagues. Finding data about soccer is easy with multiple resources but this data
set was created to attempt to collate all the different information into a single
database so as to enable data analytics and machine learning. The data is updated
regularly. The original data consisted of 201 rows which can be seen at the
[Kaggle dataset website](https://www.kaggle.com/hugomathien/soccer). We take a 
small subset of the available tables and attributes which are of interest to us.
The description for these can be found [here](docs/data_description.html).

### Data Pecularities

***

## Data Cleaning {.tabset}

### 1. Importing the data
```{r, warning = F, cache = T}
# Download the file to temp folder
temp <- tempfile()
download.file("https://kaggle2.blob.core.windows.net/datasets/63/589/database.sqlite.zip?sv=2015-12-11&sr=b&sig=btA9gYGczWR6SFFq6rG84Eok8jazkwa8JNwfYKAGtAc%3D&se=2016-12-12T16%3A25%3A12Z&sp=r",
              temp, method = 'curl')

# Connect to the sqlite file
con = dbConnect(RSQLite::SQLite(), dbname = unzip(temp))
# get a list of all tables
dbListTables(con)

# Read all tables
myQuery <- dbSendQuery(con, "SELECT * FROM Player")
Player <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Country")
Country <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM League")
League <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Player_Attributes")
Player_Attributes <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Match")
Match <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Team_Attributes")
Team_Attributes <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Team")
Team <- dbFetch(myQuery, n = -1)
# Clean up the environment
rm(temp, con, myQuery)
```

### 2. EPL games

We would like to filter out matches for only the England's top flight competition
EPL (English Premier League)

```{r}
# Find the country id for England's Premier League
England_id <- League %>% 
  filter(name == "England Premier League") %>%
  select(id) %>%
  as.double()

# Filter the matches data for only England
Match <- Match %>%
  as_tibble %>%
  filter(league_id == England_id)
```

### 3. Fixing data types

* The date column is a string, it needs to be converted to date type
* The stage and season column should be converted to factor variables
* The result column needs to be redefined with the help of a function `getResult` as:  

    ```
      if home_goal > away_goal : home
      else if home_goal == away_goal : draw
      else away
    ```
* We will also select a subset of the columns since not all of them are of interest to us
  
```{r}
# result column - this needs to be redefined as a column using the following function
getResult <- function(home_goal, away_goal){
  return (ifelse(home_goal > away_goal,
                 'home',
                 (ifelse(home_goal == away_goal,
                         'draw', 'away'))
  ))
}
```

```{r}

# Match table has:
# season column - this should be a factor variable
# date column is a string - this should be a date variable
# stage column - this should be a factor variable
# we will then select only the columns that are of interest to us

tidy_match <- Match %>%
  mutate(season = (season %>% factor(levels = unique(season))),
         date = as.Date(date),
         stage = (stage %>% factor(levels = c(1:38))),
         result =  getResult(home_team_goal, away_team_goal)) %>% 
  mutate(result = factor(result, levels = c("home", "draw", "away"))) %>%
  select(matches("id|season|stage|date|result|team_goal|goal"))

```

### 4. Parsing XML

The goal column in the data frame is in xml form we will use the `xmlToDataFrame` function from the `XML` package __to convert this into a dataframe and will then bind all the dataframes into a single one using `bind_rows`__. This dataframe will provide us all the details of the goals scored in the match. The columns of this data frame are described [here](docs/data_description.html).

```{r, warning = F, message = F}
convertXMLtoDFAndAddMatchID <- function(goal_xml, match_id) {
  df <- xmlToDataFrame(goal_xml)
  if(nrow(df) > 0){
    return(cbind(df, match_id = match_id))
  }
  return(df)
}
```

__We will also recode the goal type so that the name is more descriptive__

```{r, warning = F, message = F}
Goals <- tidy_match$goal %>% 
  map2(tidy_match$match_api_id, convertXMLtoDFAndAddMatchID) %>%
  bind_rows() %>%
  as_tibble %>%
  transmute(
    type = fct_recode(comment,
      "normal" = "n",
      "penalty" = "p",
      "disallowed goal" = "dg",
      "own goal" = "o",
      "penalty saved" = "npm",
      "penalty missed" = "psm",
      "retake penalty" = "rp"
    ),
    elapsed = as.integer(elapsed),
    player1 = as.integer(player1),
    player2 = as.integer(player2),
    subtype = factor(subtype, levels = unique(subtype)),
    team = as.integer(team),
    match_id = match_id
  )
```


### 5. Quick look at the data

We can now go ahead and delete the goal column from tidy_match and look at the two
primary data frames we will be using.

```{r}
tidy_match <- tidy_match %>% select(-goal)

tidy_match

Goals
```

### 6. Summary





