---
title: "Exploring European Soccer Database"
author: Sahil Thapar
output: html_document
---

## Synopsis
This R Markdown Notebook is my project prosposal for the [Data Wrangling with R](https://uc-r.github.io/data_wrangling/syllabus) 
I'm looking to explore European Soccer Database using dplyr and ggplot2 to predict the result of games using historical data. 

## Data Source

The data used is provided through [Kaggle datasets](https://www.kaggle.com/hugomathien/soccer).

## About the data

The data gives information on matches played across 11 different leagues.

__How leagues work?__

* Each league consists of 20 teams that play 2 games (one at home ground, one at away ground) against every other team in a round-robin fashion i.e. every team plays a total of 38 games in a season.  
* Every win is awarded with 3 points, draw with 1 point and no points are awarded for a loss.  
* The team with the most points at the end of the season wins the league.  
* The bottom 3 teams are replaced by the top-3 finishing teams from the second-tier league.  

The data consists of multiple tables. You can find a full description of variables being used [here](data_description.html).

## Packages Required

```{r, warnings = F, message = F}
library("RSQLite")
library("dplyr")
library("readr")
library("tidyverse")
library("tibble")
library('stringr')
library('lubridate')
```

## Importing the data
```{r, cache = TRUE}

temp <- tempfile()
download.file("https://kaggle2.blob.core.windows.net/datasets/63/589/database.sqlite.zip",
              temp, method = 'curl')

# Extract to get the sqlite file
data <- unzip(temp)
filename <- str_c(temp, str_sub(data, 2))

# Connect to the sqlite file
con = dbConnect(RSQLite::SQLite(), dbname = unzip(temp))
# get a list of all tables
dbListTables(con)

# Read all tables
myQuery <- dbSendQuery(con, "SELECT * FROM Player")
Player <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Country")
Country <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM League")
League <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Player_Attributes")
Player_Attributes <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Match")
Match <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Team_Attributes")
Team_Attributes <- dbFetch(myQuery, n = -1)
myQuery <- dbSendQuery(con, "SELECT * FROM Team")
Team <- dbFetch(myQuery, n = -1)
# Clean up the environment
rm(temp, filename, data, con, myQuery)
```


## Data Cleaning
```{r}
# Find the country id for England's Premier League
England_id <- League %>% 
  filter(name == "England Premier League") %>%
  select(id) %>%
  as.double()

# Filter the matches data for only England
Match <- Match %>%
  as_tibble %>%
  filter(league_id == England_id)

# Match table has:
# season column - this should be a factor variable
# date column is a string - this should be a date variable
# stage column - this should be a factor variable
# result column - this new column is defined as:
  # if home_goal > away_goal : home
  # else if home_goal == away_goal : draw
  # else away

getResult <- function(home_goal, away_goal){
  return (ifelse(home_goal > away_goal,
                 'home',
                 (ifelse(home_goal == away_goal,
                         'draw', 'away'))
  ))
}

tidy_Match <- Match %>%
  mutate(season = (season %>% factor(levels = unique(season))),
         date = as.Date(date),
         stage = (stage %>% factor(levels = c(1:38))),
         result =  getResult(home_team_goal, away_team_goal)) %>% 
  mutate(result = factor(result, levels = c("home", "draw", "away"))) %>%
  select(matches("id|season|stage|date|result|team_goal|goal"))

```


## Planned analysis

* Come up with a base model for predicting result
* Develop a naive model for predicting result
* Find most successful teams - overall and grouped by season
* Find most prolific teams, players
* Try developing a better model to predict results
* Develop Shiny app to explore team and player stats.

## Current challenges
The goal field in the Match table is in xml form. Extracting useful information from this is not simple.
