---
title: "Exploring European Soccer Database"
author: Sahil Thapar
output:
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "./")
```
## Introduction {.tabset}

### Why Football ?

Football (Soccer) is the most watched and played sport in the world. The [2014 FIFA
World Cup final was viewed by about 3.2 billion people](http://www.fifa.com/worldcup/news/y=2015/m=12/news=2014-fifa-world-cuptm-reached-3-2-billion-viewers-one-billion-watched--2745519.html).
European countries have the most followed club football leagues across the world and the English Premier League (EPL) is the most popular one.  

__How leagues work?__

* Each league consists of 20 teams that play 2 games (one at home ground, one at away ground) against every other team in a round-robin fashion i.e. every team plays a total of 38 games in a season.  
* Every win is awarded with 3 points, draw with 1 point and no points are awarded for a loss.  
* The team with the most points at the end of the season wins the league.  
* The bottom 3 teams are replaced by the top-3 finishing teams from the second-tier league.  

Using historical data, I would like to predict the result of games. Prediction of
results is not only advantageous for betting, it can also give you bragging rights
over your friends in [fantasy football](https://fantasy.premierleague.com/).  

I will employ `dplyr` and other [libraries]() to clean up data and build some simple
models to predict the result. The result for every match can be one of the following three

* __Home__: Home team wins
* __Away__: Away team wins
* __Draw__: The game ends in a draw

The data consists of multiple tables. You can find a full description of variables being used [here](data_description.html).

### Insights


### Packages Required

```{r, warnings = F, message = F}
library("knitr") # To enable usage of R code chunks from ecternal file in the Rmd
library("RSQLite") # Connecting to RSQLite
library("dplyr")   # Working 
library("readr")   # Reading and writing files
library("tidyverse") # For all the tidying functions from the tidy universe
library("tibble")    # For working with tibbles
library('stringr')   # Working with strings and regex
library('lubridate') # Working with dates
read_chunk("code/data_collection.R")
```


### Planned analysis

* Come up with a base model for predicting result
* Develop a naive model for predicting result
* Find most successful teams - overall and grouped by season
* Find most prolific teams, players
* Try developing a better model to predict results
* Develop Shiny app to explore team and player stats.

## Current challenges
The goal field in the Match table is in xml form. Extracting useful information from this is not simple.
